Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-02 appendix.tex
first complete revision after barker & latter

For the radially local problem, we eliminate $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
Eq. \ref{lin_vz}  (with $\Gamma=1$), but now retain the vertical dependence
of $D$ using Eq. \ref{dkappa2}, to obtain  
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low frequency, nearly Keplerian
approximations with the replacement $D\to \Omega_k^2$ gives, in terms
of dimensionless variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
proportional to $\epsilon^2$. For a thin disk ($\epsilon\ll1$) with
$|q|=O(1)$, as considered in this work, neglecting these terms has no
qualitative effect on the discussion in \S\ref{analytic_relax}.

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize some results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$), and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
from Eq. \ref{lin_vz}. For isothermal perturbations it is in fact
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w}. In this case, we shall not yet make the low frequency
approximation, but first make the nearly Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii \epsilon q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

@@

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\sigma} = \hat{\omega} +
% \ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\sigma}$ and
$\hat{\nu}=\imag\hat{\sigma}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|). 
\end{align}
Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), it
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered.

@@

\subsection{Explicit solutions in the thin-disk limit}\label{iso_explicit}
Here we assume a thin disk ($\epsilon\ll1$) so that $\ln\rho\simeq
-\zhat^2/2$ and $f(\zhat)\simeq \zhat$. However, we do not assume 
low frequency from the onset. Then Eq. \ref{iso_ode0} becomes 
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \frac{\ii q\epsilon \hat{k}}{1-\hat{\sigma}^2}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W = 0.
\end{align}
We remark that taking the low-frequency limit and considering
$\hat{k}^2\gg 1$, Eq. \ref{iso_ode3} becomes equivalent to Eq. 39 in
\cite{nelson13} or Eq. 28 in \cite{barker15}, although we have taken a
different route.

@@

\subsubsection{Instability with vertical shear}
The VSI corresponds to unstable inertial waves \citep{barker15}. This
is readily obtained for large $\khat^2$ by balancing
the last two terms of Eq. \ref{simple_growth} to give the low
frequency branch. Then 
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right), \label{simple_growth2}
\end{align}
which is equivalent to Eq. 34 of \cite{barker15} in the limit
$\khat^2\gg L$. This signifies instability for
$L\geq1$ since we can choose the sign 
of the square root to make $\hat{\nu} = \imag\hat{\sigma}>0$.  These
are the low frequency unstable modes seen in
Fig. \ref{compare_modes_iso_kx10} and
Fig. \ref{compare_modes_iso_kx30}, for which $\nu\propto\omega$.

>>>>

For the radially local problem, we eliminate $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
Eq. \ref{lin_vz}  (with $\Gamma=1$), but now retain the vertical dependence
of $D$ using Eq. \ref{dkappa2}, to obtain  
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low frequency, Keplerian
approximations with the replacement $D\to \Omega_k^2$ gives, in terms
of dimensionless variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
proportional to $\epsilon^2$. For a thin disk ($\epsilon\ll1$) with
$|q|=O(1)$, as considered in this work, neglecting these terms has no
qualitative effect on the discussion in \S\ref{analytic_relax}.

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$), and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
from Eq. \ref{lin_vz}. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w}. In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii \epsilon q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

@@

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\sigma} = \hat{\omega} +
% \ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\sigma}$ and
$\hat{\nu}=\imag\hat{\sigma}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|) < \frac{\epsilon |q|}{2}\mathrm{max}(|f|) . 
\end{align}
Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), it
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered.

@@

\subsection{Explicit solutions in the thin-disk limit}\label{iso_explicit}
Here we assume a thin disk ($\epsilon\ll1$) so that $\ln\rho\simeq
-\zhat^2/2$ and $f(\zhat)\simeq \zhat$. However, we do not assume 
low frequency from the onset. Then Eq. \ref{iso_ode0} becomes 
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \frac{\ii q\epsilon \hat{k}}{1-\hat{\sigma}^2}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W = 0.
\end{align}
We remark that taking the low frequency limit and considering
$\hat{k}^2\gg 1$, Eq. \ref{iso_ode3} becomes equivalent to Eq. 39 in
\cite{nelson13} or Eq. 28 in \cite{barker15}, although we have taken a
different route.

@@

\subsubsection{Instability with vertical shear}
The VSI corresponds to unstable inertial waves. This
is readily obtained for large $\khat^2$ by balancing
the last two terms of Eq. \ref{simple_growth} to give the low
frequency branch. Then 
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right), \label{simple_growth2}
\end{align}
which is equivalent to Eq. 34 of \cite{barker15} in the limit
$\khat^2\gg L$. This signifies instability for
$L\geq1$ since we can choose the sign 
of the square root to make $\hat{\nu} = \imag\hat{\sigma}>0$.  These
are the low frequency unstable modes seen in
Fig. \ref{compare_modes_iso_kx10} and
Fig. \ref{compare_modes_iso_kx30}, for which $\nu\propto|\omega|$.
